Plug-and-play dual-tree algorithm runtime analysis

### 2014-11-20 bounds-files/empirical.tex
Changes for today.

>>>>

\begin{table}

\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
\hline
{\bf Dataset} & {\bf $n$} & {\bf $d$} & {\bf $c$} & {\bf $i_t$} & {\bf NNS} & {\bf KDE} & {\bf RS} \\
\hline
cloud & 2048 & 10 & 178.5 & \\
winequality & 6497 & 11 & 4133.0 & \\
birch3 & 100000 & 2 & 1866 & \\
corel & 37749 & 32 & 23952 & \\
covertype & 581012 & 54 & \\
isolet & 7797 & 617 & 7797 & \\
miniboone & 130064 & 50 & 130064 & \\
phy & 150000 & 78 & 27731 & \\
bio & \\
lcdm & 1808174 & 3 & \\
yp-msd & 515345 & 90 & 515243 & \\
mnist & 70000 & 784 & 70000 & \\
randu & 1000000 & 10 & \\
power & 2075260 & 7 & \\
\end{tabular}
\end{table}

### 2014-11-20 bounds-files/sec03-approx-kde.tex
Remove legacy \nu.
Changes for today.

There are two simple ways to approximate $f^*(p_q)$: {\it absolute value
approximation} and {\it relative value approximation}.  Absolute value
approximation requires that each density estimate $f(p_q)$ is within $\epsilon$
of the true estimate $f^*(p_q)$:

@@

In the following two subsections, we will separately consider both the absolute
value approximation scheme and the relative value approximation scheme.

@@

A tree-independent algorithm for solving approximate kernel density estimation
with absolute value approximation
is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
correctness proof can be found in][]{curtin2013tree}.  This function
assumes a shift-invariant kernel $\mathcal{K}(p_i, p_j) = \mathcal{K}(d( p_i,
p_j ))$.  The list $f_p$ holds partial kernel density estimates for
each query point, and the list $f_n$ holds partial kernel density
estimates for each query node.  At the beginning of the dual-tree traversal, the
lists $f_p$ and $f_n$, which are both of size $O(N)$, are each
initialized to 0.

@@

    \medskip
    \STATE $f_p(p_q) \gets f_p(p_q) + \mathcal{K}(p_q, p_r)$
    \RETURN $\mathcal{K}(p_q, p_r)$
  \end{algorithmic}

@@

    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right) /\;2$
      \RETURN $\infty$
    \ENDIF

@@

\begin{thm}
Assume that $\mathcal{K}(p_i, p_j) = \mathcal{K}\left(d(p_i, p_j)\right)$ is a
shift-invariant kernel with some bandwidth $h > 0$, where $\mathcal{K}(x)$ is
smooth, non-negative, monotonically decreasing, and concave for $x \in [0, h]$
and convex for $x \in (h, \infty)$. Then, given a query set $S_q$ and a
reference set $S_r$ with
expansion constant $c_r$, and using the
approximate kernel density estimation \texttt{BaseCase()} and \texttt{Score()}
as given in Algorithms \ref{alg:kde_base_case} and \ref{alg:kde_score},
respectively, with the traversal given in Algorithm \ref{alg:cover-tree-dual},
the running time of approximate kernel density estimation for some $\epsilon$
error parameter is bounded by
%\begin{equation}
$O(c_r^{8 + \log_2 \zeta} (N + \theta))$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$ and
$\theta$ as in Theorem \ref{thm:ct-runtime}.

@@

Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to the nature of
$\mathcal{K}(\cdot)$, the maximum value of $\mathcal{K}'(\cdot)$ is
$\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  Further,
$\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le (-\epsilon /
\mathcal{K}'(h))$.  Note that

@@

Therefore, $|R^*| = \emptyset$ when
%\begin{eqnarray}
$2^{s_r^{\max} + 2} \le -\epsilon / \mathcal{K}'(h)$, or when
%\begin{equation}
$s_r^{\max} \le \log_2( -\epsilon / \mathcal{K}'(h) ) - 2$.
%\end{equation}
%
Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
\mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
alternately stated, this means that
%\begin{equation}
$d(p_q, p_r) < \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}$.
%\end{equation}
Therefore, bounding the number of points in the set
%\begin{equation}
$B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$
%\end{equation}
is sufficient to bound $|R^*|$.  For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \log_2 \omega}$.

@@

The value $\omega$ is maximized when $s_r^{\max}$ is minimized.  Using the lower
bound on $s_r^{\max}$, $\omega$ is bounded as
%
%\begin{equation}
%$\omega = \frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1} + 1} \le
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} = -2 \mathcal{K}'(h)
%\mathcal{K}^{-1}(\epsilon) \epsilon^{-1}.
%\end{equation}
$\omega = -2 \mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$.
%
Finally, with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon)
\epsilon^{-1}$, we are able to conclude that $|R^*| \le c_r^{3 + \log_2 (2
\zeta)} = c_r^{4 + \log_2 \zeta}$.  Therefore, the entire dual-tree traversal
takes $O(c_r^{8 + \log_2 \zeta} \nu N)$ time.

@@

>>>>

In the case of an infinite-tailed kernel $\mathcal{K}(\cdot, \cdot)$, the
computation cannot be accelerated; thus, attention has turned towards tractable
approximation schemes.  Two simple schemes for the approximation of $f^*(p_q)$
are well-known: {\it absolute value approximation} and {\it relative value
approximation}.  Absolute value approximation requires that each density
estimate $f(p_q)$ is within $\epsilon$ of the true estimate $f^*(p_q)$:

@@

In the following two subsections, we will separately consider both the absolute
value approximation scheme and the relative value approximation scheme, under
the assumption of a shift-invariant kernel $\mathcal{K}(p_q, p_r) =
\mathcal{K}(\| p_q - p_r \|)$ which is monotonically decreasing and
non-negative.  In addition, we assume that there exists some bandwidth $h$ such
that $\mathcal{K}(d)$ must be concave for $d \in [0, h]$ and convex for $d \in
[h, \infty)$.  This assumption implies that the magnitude of the derivative
$|\mathcal{K}'(d)|$ is maximized at $d = h$.  These are not restrictive
assumptions; most standard kernels fall into this class, including the Gaussian,
Laplacian and Epanechnikov kernels.

@@

A tree-independent algorithm for solving approximate kernel density estimation
with absolute value approximation under the previous assumptions on th kernel
is given as a \texttt{BaseCase()} function in Algorithm \ref{alg:kde_base_case}
and a \texttt{Score()} function in Algorithm \ref{alg:kde_score}  \citep[a
correctness proof can be found in][]{curtin2013tree}.  The list $f_p$ holds
partial kernel density estimates for each query point, and the list $f_n$ holds
partial kernel density estimates for each query node.  At the beginning of the
dual-tree traversal, the lists $f_p$ and $f_n$, which are both of size $O(N)$,
are each initialized to 0.  As the traversal proceeds, node combinations are
pruned if the difference between the maximum kernel value
$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r))$ and the minimum kernel
value $\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))$ is sufficiently
small (line \ref{alg:ava-kde-prune}).  If the node combination can be pruned,
then the partial node estimate is updated (line \ref{alg:ava-kde-update}).  When
node combinations cannot be pruned, \texttt{BaseCase()} may be called, which
simply updates the partial point estimate with the exact kernel evaluation (line
\ref{alg:kde-bc-update}).

@@

    \medskip
    \STATE $f_p(p_q) \gets f_p(p_q) + \mathcal{K}(p_q, p_r)$
\ref{alg:kde-bc-update}
    \RETURN $\mathcal{K}(p_q, p_r)$
  \end{algorithmic}

@@

    \IF{$\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) -
\mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r)) < \epsilon$}
\label{alg:ava-kde-prune}
      \STATE $f_n(\mathscr{N}_q) \gets f_n(\mathscr{N}_q) + |
\mathscr{D}^p(\mathscr{N}_r) | \left(\mathcal{K}(d_{\min}(\mathscr{N}_q,
\mathscr{N}_r)) + \mathcal{K}(d_{\max}(\mathscr{N}_q, \mathscr{N}_r))\right)
/\;2$ \label{alg:ava-kde-update}
      \RETURN $\infty$
    \ENDIF

@@

\begin{thm}
Assume that $\mathcal{K}(\cdot, \cdot)$ is a kernel satisfying the assumptions
of the previous section.  Then, given a query set $S_q$ and a reference set
$S_r$ with expansion constant $c_r$, and using the approximate kernel density
estimation \texttt{BaseCase()} and \texttt{Score()} as given in Algorithms
\ref{alg:kde_base_case} and \ref{alg:kde_score}, respectively, with the
traversal given in Algorithm \ref{alg:cover-tree-dual}, the running time of
approximate kernel density estimation for some $\epsilon$ error parameter is
bounded by
%\begin{equation}
$O(c_r^{8 + \log_2 \zeta} (N + \theta))$
%\end{equation}
where $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$ and
$\theta$ as in Theorem \ref{thm:ct-runtime}.

@@

Thus, we are interested in the maximum possible value $\mathcal{K}(a) -
\mathcal{K}(b)$ for a fixed value of $b - a > 0$.  Due to our assumptions, the
maximum value of $\mathcal{K}'(\cdot)$ is
$\mathcal{K}'(h)$; therefore, the maximum possible value of $\mathcal{K}(a) -
\mathcal{K}(b)$ is when the interval $[a, b]$ is centered on $h$.  Further,
$\mathcal{K}(a) - \mathcal{K}(b) \le \epsilon$ when $(b - a) \le (-\epsilon /
\mathcal{K}'(h))$.  Note that

@@

Therefore, $R^* = \emptyset$ when
%\begin{eqnarray}
$2^{s_r^{\max} + 2} \le -\epsilon / \mathcal{K}'(h)$, or when
%\begin{equation}
$s_r^{\max} \le \log_2( -\epsilon / \mathcal{K}'(h) ) - 2$.
%\end{equation}
%
Consider, then, the case when $s_r^{\max} > \log_2( -\epsilon /
\mathcal{K}'(h) ) - 2$.  It is trivial to see that for any $\mathscr{N}_r \in
R^*$, $\mathcal{K}(d_{\min}(\mathscr{N}_q, \mathscr{N}_r)) > \epsilon$;
alternately stated, this means that
%\begin{equation}
$d(p_q, p_r) < \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}$.
%\end{equation}
Therefore, bounding the number of points in the set
%\begin{equation}
$B_{S_r}(p_q, \mathcal{K}^{-1}(\epsilon) + 2^{s_r^{\max} + 1}) \cap
C_{s_r^{\max}}$
%\end{equation}
is sufficient to bound $|R^*|$.  For notational convenience, define $\omega = \mathcal{K}^{-1}(\epsilon) /
2^{s_r^{\max} + 1} + 1$.  Using Lemma \ref{lem:packing} with $\delta =
2^{s_r^{\max}}$ and $\rho = 2 \omega$ gives $|R^*| = c_r^{3 + \log_2 \omega}$.

@@

The value $\omega$ is maximized when $s_r^{\max}$ is minimized.  Using the lower
bound on $s_r^{\max}$, $\omega$ is bounded as
%
%\begin{equation}
%$\omega = \frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1} + 1} \le
%\frac{\mathcal{K}^{-1}(\epsilon)}{2^{s_r^{\max} + 1}} = -2 \mathcal{K}'(h)
%\mathcal{K}^{-1}(\epsilon) \epsilon^{-1}.
%\end{equation}
$\omega = -2 \mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon) \epsilon^{-1}$.
%
Finally, with $\zeta = -\mathcal{K}'(h) \mathcal{K}^{-1}(\epsilon)
\epsilon^{-1}$, we are able to conclude that $|R^*| \le c_r^{3 + \log_2 (2
\zeta)} = c_r^{4 + \log_2 \zeta}$.  Therefore, the entire dual-tree traversal
takes $O(c_r^{8 + \log_2 \zeta} (N + \theta))$ time.

@@

To understand the lack of dependence on kernel bandwidth more intuitively,
consider that as the kernel bandwidth increases, two things happen: {\it (a)}
the reference set $R$ becomes empty at larger scales, and {\it (b)}
$\mathcal{K}^{-1}(\epsilon)$ grows, allowing less pruning at higher levels.
These effects are opposite, and for the Gaussian kernel they cancel each other
out, giving the same bound regardless of bandwidth.

Other kernels do not exhibit this effect.  Consider the Laplacian kernel:
